### 2023-09-10 proofs/tex/norm-lemmas.tex
Add deterinicity lemmas, get rid of (??)
Add "straightforward" proofs

\begin{lemma}
  \label{lemma:equiv-fv}
  Set of free variables is invariant under equivalence.
  \begin{itemize}
  \item[$-$] If $[[iN ≈ iM]]$ then $[[fv iN]] \equiv [[fv iM]]$ (as sets)
  \item[$+$] If $[[iP ≈ iQ]]$ then $[[fv iP]] \equiv [[fv iQ]]$ (as sets)
  \end{itemize}
\end{lemma}
\begin{proof}
  Straightforward mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$
\end{proof}

\begin{lemma}
  \label{lemma:fv-nf}
  Free variables are not changed by the normalization
  \begin{itemize}
  \item[$-$] $[[fv iN]] \equiv [[fv nf(iN)]]$
  \item[$+$] $[[fv iP]] \equiv [[fv nf(iP)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  By straightforward induction on $[[iN]]$ and mutually on $[[iP]]$.
\end{proof}

>>>>

\begin{observation}[Normalization is deterministic]
  \label{obs:norm-deterministic}
  If $[[nf(iN) = iM]]$ and $[[nf(iN) = iM']]$ then $[[iM = iM']]$.
  If $[[nf(iP) = iQ]]$ and $[[nf(iP) = iQ']]$ then $[[iQ = iQ']]$.
  This way, we can use normalization as a function.
\end{observation}
\begin{proof}
  By straightforward induction using \cref{obs:ord-deterministic}.
\end{proof}

\begin{lemma}
  \label{lemma:equiv-fv}
  Set of free variables is invariant under equivalence.
  \begin{itemize}
  \item[$-$] If $[[iN ≈ iM]]$ then $[[fv iN = fv iM]]$ (as sets)
  \item[$+$] If $[[iP ≈ iQ]]$ then $[[fv iP = fv iQ]]$ (as sets)
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$
  The base cases 
  (\ruleref{\ottdruleEOneNVarLabel} and \ruleref{\ottdruleEOnePVarLabel})
  are trivial.
  So are \ruleref{\ottdruleEOneShiftULabel}, 
  \ruleref{\ottdruleEOneShiftDLabel}, and
  \ruleref{\ottdruleEOneArrowLabel}, where
  the required property follows from the induction hypothesis.

  Let us consider the case when the equivalence is formed by
  \ruleref{\ottdruleEOneForallLabel}, 
  that is the equivalence has shape 
  $[[∀pas.iN ≈ ∀pbs.iM]]$, 
  and by inversion,
  there is a bijection 
  $[[mu : ({pbs} ∩ fv iM) ↔ ({pas} ∩ fv iN)]]$
  such that $[[iN ≈ [mu] iM]]$, which 
  by the induction hypothesis means
  $[[ fv iN = fv [mu]iM ]] = [[ [mu] fv iM]] $.

  Let us ensure by alpha-equivalence that 
  $[[pas]]$ is disjoint from $[[fv iM]]$.
  Then $[[ (fv ∀pbs.iM) \ {pas}]] = [[fv ∀pbs.iM]]$.
  Then we apply the following chain of equalities:
  $[[fv ∀pas.iN = fv iN \ {pas}]] = 
  [[ ([mu] fv iM) \ {pas}]] = 
  [[ [mu] (fv ∀pbs.iM ∪ ({pbs} ∩ fv iM)) \ {pas}]] = 
  [[ ([mu] fv ∀pbs.iM ∪ [mu]({pbs} ∩ fv iM)) \ {pas}]] = 
  [[ ([mu] fv ∀pbs.iM) \ {pas}]] = 
  [[ (fv ∀pbs.iM) \ {pas}]] = 
  [[  fv ∀pbs.iM ]]$.

  Symmetrically, we prove the case when the equivalence is formed by
  \ruleref{\ottdruleEOneExistsLabel}.
\end{proof}

\begin{lemma}
  \label{lemma:fv-nf}
  Free variables are not changed by the normalization
  \begin{itemize}
  \item[$-$] $[[fv iN = fv nf(iN)]]$
  \item[$+$] $[[fv iP = fv nf(iP)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  By mutual induction on $[[iN]]$ and $[[iP]]$.
  The base cases 
  (\ruleref{\ottdruleNrmNVarLabel} and \ruleref{\ottdruleNrmPVarLabel})
  are trivial; the congruent cases
  (\ruleref{\ottdruleNrmShiftULabel},
  \ruleref{\ottdruleNrmShiftDLabel}, and
  \ruleref{\ottdruleNrmArrowLabel}) are proved by the induction hypothesis.

  Let us consider the case when the term is formed by $[[∀]]$,
  that is the normalization judgement has shape 
  $[[nf(∀pas.iN) = ∀pas'.iN']]$,
  where by inversion $[[nf(iN) = iN']]$
  and $[[ord {pas} in iN' = pas']]$.
  By the induction hypothesis, $[[fv iN = fv iN']]$.
  Since $[[fv(∀pas.iN) = fv iN \ {pas}]]$,
  and $[[fv(∀pas'.iN') = fv iN' \ {pas'}]]$,
  it is left to show that $[[fv iN \ {pas} = fv iN \ {pas'}]]$.
  By \cref{lemma:ord-completeness}, 
  $[[{pas'}]] = [[{pas} ∩ fv iN']] = [[{pas} ∩ fv iN]]$.
  Then
  $[[fv iN \ {pas} = fv iN \ ({pas} ∪ fv iN)]]$ by
  set-theoretic properties, and
  thus, $[[fv iN \ {pas} = fv iN \ {pas'}]]$.

  The case when the term is positive and formed by $[[∃]]$ is symmetric.
\end{proof}

### 2023-09-10 proofs/tex/subst-lemmas.tex
Add "straightforward" proofs

\begin{lemma}[Free variables of a variable image]
  \label{lemma:subst-fv-image}
  Suppose that $[[σ]]$ is an arbitrary substitution, 
  Then 
  \begin{itemize}
    \item [$+$] if $[[α± ∊ fv(iP)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iP) ]]$,
    \item [$-$] if $[[α± ∊ fv(iN)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iN) ]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  By straightforward mutual induction on $[[iP]]$ and $[[iN]]$.
\end{proof}

>>>>

\begin{lemma}[Free variables of a variable image]
  \label{lemma:subst-fv-image}
  Suppose that $[[σ]]$ is an arbitrary substitution, 
  Then 
  \begin{itemize}
    \item [$+$] if $[[α± ∊ fv(iP)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iP) ]]$,
    \item [$-$] if $[[α± ∊ fv(iN)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iN) ]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  By mutual induction on $[[iP]]$ and $[[iN]]$.
  The base cases (when $[[iP]]$ or $[[iN]]$ is a variable) 
  are trivial, since then $[[α± ∊ fv(iP)]]$ means 
  $[[α± = iP]]$ (and symmetrically for $[[iN]]$).
  The congruent cases 
  (when the type is formed by $[[↓]]$, $[[↑]]$, or $[[→]]$)
  hold since $[[α±]]$ occurs in type means that it occurs
  in one of its parts, to which we apply the induction hypothesis.

  Let us suppose that te type is $[[∃nas.iQ]]$.
  Then $[[α± ∊ fv(∃nas.iQ)]]$ means
  $[[α± ∊ fv(iQ)]]$ and $[[α± ∉ {nas}]]$.
  Then by the induction hypothesis, 
  $[[ fv([σ]α±) ⊆ fv([σ]iQ) ]]$,
  and it is left to notice that $[[ fv([σ]α±) ∩ {nas} = ∅ ]]$,
  which we can ensure by alpha-equivalence.
\end{proof}
